---
title: "Draft: Data Documentation"
---
https://www.spc.noaa.gov/wcm/data/SPC_severe_database_description.pdf
```{r, message = FALSE}
library(tidyverse)

tornado_full_data <- read.csv("1950-2024_actual_tornadoes.csv")

tornado_data <- tornado_full_data |>
  select(yr, date, st, mag, slat, slon, elat, elon, len, wid, ns, sn, sg)

tornado_data_counted_once <- tornado_data |>
  filter(sn == 0) |>
  select(-ns,-sn,-sg) |>
  arrange(date)
```

```{r, echo = FALSE}
tornado_vars <- data.frame(
  Variable = c("yr", "date", "st", "mag", "slat", "slon", "elat", "elon", "len", "wid"),
  Description = c(
    "Year Tornado Occurred",
    "Exact Date Tornado Occurred",
    "State Abbreviation Where Tornado First Touched Down",
    "EF Scale Magnitude of the Tornado",
    "Latitude Where the Tornado First Touched Down",
    "Longitude Where the Tornado First Touched Down",
    "Latitude Where the Tornado Ended",
    "Longitude Where the Tornado Ended",
    "Tornado Path Length (Miles)",
    "Tornado Path Width (Yards)"
  ),
  Type = c("Integer", "Date", "Categorical", "Integer", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric"),
  Possible_Values = c("1950-2024", "01-01-1950 to 12/31/2024", "Any US State", "0-5 or -9 Meaning Unknown", "US Latitudes ~ 29 to 50", "US Longitudes ~ -125 to -70", "US Latitudes ~ 29 to 50", "US Longitudes ~ -125 to -70", "0-250 Miles", "0-3500 Yards")
)
```

```{r, echo = FALSE}
library(gt)

tornado_vars %>%
  gt() %>%
  tab_header(
    title = "Tornado Data Variable Descriptions"
  ) %>%
  cols_label(
    Variable = "Variable Name",
    Description = "Description",
    Type = "Type",
    Possible_Values = "Possible Values"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = cell_borders(
      sides = "all",
      color = "black",
      weight = px(1)
    ),
    locations = cells_body(everything())
  ) %>%
  tab_style(
    style = cell_borders(
      sides = "all",
      color = "black",
      weight = px(1)
    ),
    locations = cells_column_labels(everything())
  )

```




The data were collected by NOAA National Weather Prediction Service.

The data was collected simply for the purpose of historical records and public safety, so that there can be research into these tornadoes.

The data is about tornadoes, showing where they occur and how they travel.

The data was collected from 1950-2024, and shows the date each tornado occurred as a column in the dataset.

The data was collected all around the United States. This includes (potentially) all states.

Reports from storm spotters and what the NWS considers reputable sources are submitted to the Storm Data publication by National Weather Service field offices.

The data is a simple CSV file with each row representing one tornado and each column being a different descriptor of the tornado.

The NOAA is a reputable known source for these types of data so there should be no data quality issues.

Some suggested data analysis methods include probability estimation based on magnitude and/or distance travelled inland or towards the coast











