---
title: "Executive Summary"
format: html
---

This report estimates the probability of a “sharknado” by incorporating real-world constraints and combining the probabilities of the events required for such an occurrence. Using available tornado data and reasonable assumptions, the analysis evaluates the likelihood of three key components:

1)  The presence of a shark on the coast
2)  A tornado moving coastal to inland
3)  A tornado is strong enough to lift and move a shark inland

The analysis sequences the relevant probabilities to estimate the overall likelihood. A key assumption is that sharks are always present when a coastal-to-inland tornado occurs, which intentionally overestimates the probability and provides a conservative upper bound. The results indicate that, even under favorable assumptions, the probability of a sharknado is effectively zero. Only one tornado since 1950 shows the possibility of being a sharknado, an EF4 tornado from Florida in 1966.


```{r, echo = FALSE, message = FALSE, output = FALSE}
library(gt)

tornado_vars <- data.frame(
  Variable = c("yr", "date", "st", "mag", "slat", "slon", "elat", "elon", "len", "wid"),
  Description = c(
    "Year Tornado Occurred",
    "Exact Date Tornado Occurred",
    "State Abbreviation Where Tornado First Touched Down",
    "EF Scale Magnitude of the Tornado",
    "Latitude Where the Tornado First Touched Down",
    "Longitude Where the Tornado First Touched Down",
    "Latitude Where the Tornado Ended",
    "Longitude Where the Tornado Ended",
    "Tornado Path Length (Miles)",
    "Tornado Path Width (Yards)"
  ),
  Type = c("Integer", "Date", "Categorical", "Integer", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric"),
  Possible_Values = c("1950-2024", "01-01-1950 to 12/31/2024", "Any US State", "0-5 or -9 Meaning Unknown", "US Latitudes ~ 29 to 50", "US Longitudes ~ -125 to -70", "US Latitudes ~ 29 to 50", "US Longitudes ~ -125 to -70", "0-250 Miles", "0-3500 Yards")
)

tornado_vars %>%
  gt() %>%
  tab_header(
    title = "Tornado Data Variable Descriptions"
  ) %>%
  cols_label(
    Variable = "Variable Name",
    Description = "Description",
    Type = "Type",
    Possible_Values = "Possible Values"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = cell_borders(
      sides = "all",
      color = "black",
      weight = px(1)
    ),
    locations = cells_body(everything())
  ) %>%
  tab_style(
    style = cell_borders(
      sides = "all",
      color = "black",
      weight = px(1)
    ),
    locations = cells_column_labels(everything())
  )
# Data Cleaning (1 row per tornado, only important columns, dealing with 0's in elat and elon)
library(tidyverse)

tornado_full_data <- read.csv("1950-2024_actual_tornadoes.csv") |>
  filter(!st %in% c("AK", "HI", "PR", "VI"))

tornado_paths <- tornado_full_data |>
  distinct(slat, slon, elat, elon, .keep_all = TRUE)

tornado_paths <- tornado_paths |>
  select(slat, slon, elat, elon, mag)

tornado_paths <- tornado_paths |>
  mutate(end_missing = (elat == 0 | elon == 0))

ggplot(tornado_paths, aes(x = slon, y = slat)) +
  geom_point(color = "red", alpha = 0.5) +
  labs(title = "Starting Location For All Tornadoes",
       x = "Starting Longitude", y = "Starting Latitude") +
  theme_minimal()
ggplot(tornado_paths |> filter(elat == 0 | elon == 0),
       aes(x = slon, y = slat)) +
  geom_point(color = "red", alpha = 0.5) +
  labs(title = "Starting Location For Tornadoes With Missing End Coordinates",
       x = "Starting Longitude", y = "Starting Latitude") +
  theme_minimal()
library(ggtext)
ggplot(
  tornado_paths |> 
    filter(elat == 0 | elon == 0),
  aes(x = slon, y = slat)
) +
  geom_point(color = "red", alpha = 0.5) +
  facet_wrap(~ mag) +
  labs(
    title = "Starting Location For Tornadoes With Missing End Coordinates (by Magnitude)",
    x = "Starting Longitude", 
    y = "Starting Latitude"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown()
  )
# Transforming latitude and longitude into math ready quantities
library(rnaturalearth)
library(sf)

tornado_paths <- tornado_paths |>
  filter(end_missing == FALSE)

tornado_paths$start_geom_degrees <- st_as_sf(
  tornado_paths,
  coords = c("slon", "slat"),
  crs = 4269
)$geometry

tornado_paths$end_geom_degrees <- st_as_sf(
  tornado_paths,
  coords = c("elon", "elat"),
  crs = 4269
)$geometry

tornado_paths$start_geom_meters <- st_transform(
  tornado_paths$start_geom_degrees, 5070
)

tornado_paths$end_geom_meters <- st_transform(
  tornado_paths$end_geom_degrees, 5070
)

coast_m <- ne_coastline(returnclass = "sf") |> 
  st_transform(5070)
# Find distances to coast and classify each tornado as coastal_to_inland , inland_to_coastal , or other
tornado_paths <- tornado_paths |>
  mutate(
    dist_start_to_coast = st_distance(start_geom_meters, coast_m) |> apply(1, min),
    dist_end_to_coast = st_distance(end_geom_meters, coast_m) |> apply(1, min)
  )

tornado_paths <- tornado_paths |> 
  mutate(  
    movement = case_when(  
      dist_start_to_coast <= 10000 & dist_end_to_coast > dist_start_to_coast ~ "Coastal_to_Inland",  
      dist_start_to_coast > 10000 & dist_end_to_coast < dist_start_to_coast ~ "Inland_to_Coastal",  
      TRUE ~ "Other"  
    )  
  )
```

```{r, echo = FALSE}
florida_tornado <- tornado_paths |>
  filter(mag == 4 & movement == "Coastal_to_Inland")

florida <- map_data("state") |>
  filter(region == "florida")

ggplot() +
  geom_polygon(
    data = florida,
    aes(x = long, y = lat, group = group),
    fill = NA,
    color = "black",
    linewidth = 0.3
  ) +
  geom_point(
    data = florida_tornado,
    aes(x = slon, y = slat),
    color = "red",
    size = 3
  ) +
  labs(
    title = "Only Possible Sharknado Found",
    x = "Longitude",
    y = "Latitude"
  )
```

::: callout-important
The probability of a sharknado is effectively zero. As such, no adjustments to insurance products are warranted. Resources should instead focus on higher-probability catastrophic events such as hurricanes and standard tornadoes.
:::
