---
title: "Results"
format: html
---

```{r, echo = FALSE, message = FALSE, output = FALSE}
library(gt)

tornado_vars <- data.frame(
  Variable = c("yr", "date", "st", "mag", "slat", "slon", "elat", "elon", "len", "wid"),
  Description = c(
    "Year Tornado Occurred",
    "Exact Date Tornado Occurred",
    "State Abbreviation Where Tornado First Touched Down",
    "EF Scale Magnitude of the Tornado",
    "Latitude Where the Tornado First Touched Down",
    "Longitude Where the Tornado First Touched Down",
    "Latitude Where the Tornado Ended",
    "Longitude Where the Tornado Ended",
    "Tornado Path Length (Miles)",
    "Tornado Path Width (Yards)"
  ),
  Type = c("Integer", "Date", "Categorical", "Integer", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric"),
  Possible_Values = c("1950-2024", "01-01-1950 to 12/31/2024", "Any US State", "0-5 or -9 Meaning Unknown", "US Latitudes ~ 29 to 50", "US Longitudes ~ -125 to -70", "US Latitudes ~ 29 to 50", "US Longitudes ~ -125 to -70", "0-250 Miles", "0-3500 Yards")
)

tornado_vars %>%
  gt() %>%
  tab_header(
    title = "Tornado Data Variable Descriptions"
  ) %>%
  cols_label(
    Variable = "Variable Name",
    Description = "Description",
    Type = "Type",
    Possible_Values = "Possible Values"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = cell_borders(
      sides = "all",
      color = "black",
      weight = px(1)
    ),
    locations = cells_body(everything())
  ) %>%
  tab_style(
    style = cell_borders(
      sides = "all",
      color = "black",
      weight = px(1)
    ),
    locations = cells_column_labels(everything())
  )
# Data Cleaning (1 row per tornado, only important columns, dealing with 0's in elat and elon)
library(tidyverse)

tornado_full_data <- read.csv("1950-2024_actual_tornadoes.csv") |>
  filter(!st %in% c("AK", "HI", "PR", "VI"))

tornado_paths <- tornado_full_data |>
  distinct(slat, slon, elat, elon, .keep_all = TRUE)

tornado_paths <- tornado_paths |>
  select(slat, slon, elat, elon, mag)

tornado_paths <- tornado_paths |>
  mutate(end_missing = (elat == 0 | elon == 0))

ggplot(tornado_paths, aes(x = slon, y = slat)) +
  geom_point(color = "red", alpha = 0.5) +
  labs(title = "Starting Location For All Tornadoes",
       x = "Starting Longitude", y = "Starting Latitude") +
  theme_minimal()
ggplot(tornado_paths |> filter(elat == 0 | elon == 0),
       aes(x = slon, y = slat)) +
  geom_point(color = "red", alpha = 0.5) +
  labs(title = "Starting Location For Tornadoes With Missing End Coordinates",
       x = "Starting Longitude", y = "Starting Latitude") +
  theme_minimal()
library(ggtext)
ggplot(
  tornado_paths |> 
    filter(elat == 0 | elon == 0),
  aes(x = slon, y = slat)
) +
  geom_point(color = "red", alpha = 0.5) +
  facet_wrap(~ mag) +
  labs(
    title = "Starting Location For Tornadoes With Missing End Coordinates (by Magnitude)",
    x = "Starting Longitude", 
    y = "Starting Latitude"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown()
  )
# Transforming latitude and longitude into math ready quantities
library(rnaturalearth)
library(sf)

tornado_paths <- tornado_paths |>
  filter(end_missing == FALSE)

tornado_paths$start_geom_degrees <- st_as_sf(
  tornado_paths,
  coords = c("slon", "slat"),
  crs = 4269
)$geometry

tornado_paths$end_geom_degrees <- st_as_sf(
  tornado_paths,
  coords = c("elon", "elat"),
  crs = 4269
)$geometry

tornado_paths$start_geom_meters <- st_transform(
  tornado_paths$start_geom_degrees, 5070
)

tornado_paths$end_geom_meters <- st_transform(
  tornado_paths$end_geom_degrees, 5070
)

coast_m <- ne_coastline(returnclass = "sf") |> 
  st_transform(5070)
# Find distances to coast and classify each tornado as coastal_to_inland , inland_to_coastal , or other
tornado_paths <- tornado_paths |>
  mutate(
    dist_start_to_coast = st_distance(start_geom_meters, coast_m) |> apply(1, min),
    dist_end_to_coast = st_distance(end_geom_meters, coast_m) |> apply(1, min)
  )

tornado_paths <- tornado_paths |> 
  mutate(  
    movement = case_when(  
      dist_start_to_coast <= 10000 & dist_end_to_coast > dist_start_to_coast ~ "Coastal_to_Inland",  
      dist_start_to_coast > 10000 & dist_end_to_coast < dist_start_to_coast ~ "Inland_to_Coastal",  
      TRUE ~ "Other"  
    )  
  )
```

::: callout
introduce the section
:::

Now we need to compute the probabilities for each tornado type that we have classified. This can be done quite simply by dividing the number of tornadoes that we have classified to each tornado type by the total number of tornadoes. The results are summarized in the following table:

```{r, echo = FALSE}
library(knitr)
# Find probabilities of each type of tornado
number_coastal_to_inland <- tornado_paths |>
  filter(movement == "Coastal_to_Inland") |>
  nrow()

number_inland_to_coastal <- tornado_paths |>
  filter(movement == "Inland_to_Coastal") |>
  nrow()

number_other <- tornado_paths |>
  filter(movement == "Other") |>
  nrow()

total_tornadoes <- nrow(tornado_paths)

probability_coastal_to_inland <- 100*(number_coastal_to_inland/total_tornadoes)
probability_inland_to_coastal <- 100*(number_inland_to_coastal/total_tornadoes)
probability_other <- 100*(number_other/total_tornadoes)

tornado_df_for_table <- data.frame(
  category = c("Coastal to Inland", "Inland to Coastal", "Other"),
  number_of_tornadoes = c(number_coastal_to_inland, number_inland_to_coastal, number_other),
  probability = c(probability_coastal_to_inland, probability_inland_to_coastal, probability_other)
)

kable(
  transform(
    tornado_df_for_table,
    probability = sprintf("%.2f%%", probability)
  ),
  col.names = c("Tornado Type", "Number of Tornadoes", "Probability")
)


```

As shown, there are far fewer coastal-to-inland tornadoes compared with inland-to-coastal and other tornadoes. This relatively small amount reflects the strict requirement that the tornado must start within a short distance of the coast to be classified in this category.

```{r, echo = FALSE}
#Find probabilities of each magnitude of each type of tornado
options(scipen = 999)

probabilities <- tornado_paths |>
  group_by(mag, movement) |>
  summarise(count = n(), .groups = "drop") |>
  group_by(movement) |>
  mutate(magnitude_proportion_given_movement_type = count / sum(count)) |>
  ungroup() |>
  mutate(proportion_of_total_tornadoes = count / total_tornadoes)

probabilities_for_coastal_to_inland <- probabilities |>
  filter(movement == "Coastal_to_Inland") |>
  mutate(magnitude_probability_given_movement_type = 100*magnitude_proportion_given_movement_type)
```

Lastly, we estimate the probability that a tornado is strong enough to lift a shark, conditional on it being coastal-to-inland. According to @fisheriesWhiteSharkNOAA2025, great white sharks can weigh up to 4,500 lbs. Similarly, @HowMuchDoes reports that the average car weighs approximately 4,419 lbs, suggesting that great white sharks are of comparable mass to cars. From @CanCarBe2025, EF3 tornadoes are “likely to toss cars and trucks,” whereas EF4 tornadoes are “very likely to toss cars many feet.” This indicates that at least an EF4 tornado would be required, since a sharknado would require sharks not only to be lifted, but also transported a substantial distance inland rather than merely displaced. To compute this probability, we first have to define the number of tornadoes for each combination of magnitude $m$ and movement type $t$, call this $N_{m,t}$. The probability of a tornado having magnitude $m$ given movement type $t$ is the number of tornadoes with magnitude $m$ and movement type $t$ divided by the total number of tornadoes with movement type $t$.

::: callout
also mention that they have to be pulled out of water - is that possible to calculate with water being an added effect?
:::

$$
P(\text{Magnitude} = m|\text{Movement} = t) = \frac{N_{m,t}}{\sum_{m} N_{m,t}}
$$ For our specific probability we are interested in this probability where m = 4 or 5, and t = coastal to inland.

Computing this probability we get:

$$
\frac{`r probabilities_for_coastal_to_inland[6,"count"]` + 0}{`r sum(probabilities_for_coastal_to_inland$count)`} 
\approx `r round(probabilities_for_coastal_to_inland[6,"magnitude_probability_given_movement_type"], 2)`\%
\quad
(1 \text{ from magnitude 4},\ 0 \text{ from magnitude 5})
$$

::: callout
double check how the math notation is interpreting your text visually since underscores turn the following letters or symbols into a subscript. Also what is the ending probability numerically other than converging to zero?
:::

It is important to note that even this estimate represents a conservative upper bound. We assume that a shark is present along the coast whenever a coastal-to-inland tornado occurs, and that an EF4 tornado is sufficient to lift a shark from the ocean onto land. These assumptions are deliberately favorable to the occurrence of a sharknado. Relaxing these large assumptions would cause this probability to converge even closer to 0.
